---
title: "Volleyball Percentiles"
---

We'll be calculating percentiles to see how good Maryland was in service aces in the 2024 season. We'll use `mutate()` and a new function `ntile()` to calculate percentiles. You should open the VBPercentiles_Explainer.Rmd document in the explainers folder and use that as a reference.

Get started by loading the tidyverse in a code block, then load the CSV file here: https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/refs/heads/main/data/ncaa_womens_volleyball_matchstats_2024.csv" and save it to a dataframe called `matches`.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the data

```{r}
matches <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/refs/heads/main/data/ncaa_womens_volleyball_matchstats_2024.csv")
```

### Task 3: Create totals of aces for each team using `group_by()` and `summarize()`

```{r}
team_aces <- matches |> 
  group_by(team) |> 
  summarize(total_aces = sum(aces)) |> 
  arrange(desc(total_aces))
```


### Task 4: Calculate percentiles using `mutate()` and `ntile()`

```{r}
team_aces |> mutate(percentile = ntile(total_aces, 100))
```


### Task 5: Describe the percentiles

Write a couple of sentences that puts Maryland into context with the other teams in the dataset. You can do some Internet research to help with this.

Maryland has massive institutional funding, elite training facilities, top tier medical staff, and have high athletic scholarship funding that supports them in program quality scores, which puts them on a higher level compared to other teams who potentially don't have the same financial backing and resources. Being members of the Big Ten conference also automatically places them at the very top of most other teams in the nation.

When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.






