---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r} 
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}

md_cities_rates <- md_cities |>mutate( 
violent_rate_2019= (violent2019 / pop2019) * 1000,
property_rate_2020= (property2020 / pop2020) * 1000,
property_rate_2019= (property2019 / pop2019) * 1000,
violent_rate_2020= (violent2020 / pop2020) * 1000
) 

```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

ANSWER HERE: The city with the highest rate of violent crime in 2019 was Baltimore, which makes sense because although in recent times Baltimore's crime rate has gone down in recent years, it used to be seen as a hub for violent crimes like shootings, murders, and robberys. Overall I noticed there were overall higher rates of property crimes than in violent crimes in the year 2019, which makes sense. The city with the highest property crime rate in 2019 was Ocean City. This may be due to them being a City with a high number of tourists. In 2020, the city with the highest violent crime rate was Ocean City, which was surprising as I'd expect it to still be Baltimore, but the COVID-19 pandemic may have influenced this, (although the data does not specify this) The city with the highest rate of property crime in 2020 was Ocean City again, which is surprising. I notice that there are some cities in certain counties with very low populations ex. 100-1000 people and some of them reported 0 violent and property crimes. 


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> mutate(percentage_change_violent= (violent_rate_2020 - violent_rate_2019) / violent_rate_2019 * 100, percentage_change_property= (property_rate_2020 - property_rate_2019) / property_rate_2019 * 100)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

ANSWER HERE: I notice that the percentage changes show that crime rates varied significantly between Maryland cities from 2019 to 2020, with some smaller cities experiencing especially large increases or decreases. I think the outlier's are more interesting than the overall pattern because cities with very small populations can show dramatic percentage changes from only a few additional or fewer reported crimes. This suggests that while the data can reveal changes in crime rates, the more extreme percentage changes should be looked at carefully because they could just be influenced by smaller populations.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE: Ocean city recorded a sharp increase in its violent crime rate in 2020, surpassing Baltimore City to have the highest violent crime rate among Maryland cities. 

The data shows that crime rates varies across Maryland cities between 2019 and 2020. Ocean city had some of the highest rates in the data set overall. The data also shows that smaller cities can still have relatively high crime rates because even a small number of reported crimes can have a larger effected when compared with a smaller population. This data shows the importance of population size when interperting crime statistics, as raw crime totals can make big cities look worse simply because they have more people, whereas rates give you a way to compare cities of different sizes.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
