---
title: "HW7_CarAccidents"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Analyzing Daily Accident Data with Z-Scores

## Introduction

In this analysis, we'll examine daily accident data from Prince George's County to identify statistically unusual patterns that might lead to newsworthy stories. Z-scores provide a powerful statistical method to determine which days experienced an abnormal number of accidents compared to typical patterns.

A z-score tells us how many standard deviations a data point is from the mean. Generally:
- Z-scores between -2 and 2 represent typical values (about 95% of the data)
- Z-scores beyond ±2 indicate unusual values that might warrant investigation
- Z-scores beyond ±3 are very unusual (only about 0.3% of normally distributed data)

## Setup

Let's start by loading the necessary packages:

```{r load-packages}
library(tidyverse)  # For data manipulation and visualization
library(knitr)      # For nice tables
library(lubridate)  # For working with dates
```

### Task 1: Loading the Data (1 point)

Load the accident data from our CSV file at https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/daily_accidents.csv into a dataframe called `accidents`:

```{r load-data}
accidents <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/daily_accidents.csv")
```

## Data Exploration

Let's first understand our dataset structure:

```{r data-exploration}
# Get a summary of the data
summary(accidents)

# Check how many days we have in total
nrow(accidents)

# Check the date range
min(accidents$date)
max(accidents$date)
```

Now let's create a time series visualization of accidents:

```{r time-series-plot}
# Plot time series
ggplot(accidents, aes(x = date, y = total)) +
  geom_line(alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.1) +
  theme_minimal() +
  labs(title = "Daily Accidents in Prince George's County",
       x = "Date",
       y = "Number of Accidents",
       caption = "Source: Prince George's County Accident Data")
```

### Task 2 (3 points)
Looking at the time series visualization above, identify at least two noticeable patterns or anomalies. What questions would you want to investigate further based on this initial view of the data? How might these visual patterns translate into potential story angles?

ANSWER HERE : The data above shows a sharp decline in daily accidents arround 2020 and a significant increase in 2021 before it levelled off around the mid to high 20s. there are some extreme spikes and unusually low days that could be worth investigating. I'd want to know whether the 2020 decline in accidents was related to the Covid-19 pandemic and what caused it to go back up in 2021, and whether the daily fluctuations are caused by things like weather and traffic volume. These patterns could potentially lwad to a story about how the pandemic changed driving and accident rates in PG County. 

The daily data is quite noisy, so let's also look at weekly patterns:

```{r weekly-patterns}
# Add day of week 
accidents <- accidents |>
  mutate(
    day_of_week = wday(date, label = TRUE),
    month = floor_date(date, "month"),
    year = year(date)
  )

# Check accident patterns by day of week
day_of_week_summary <- accidents |>
  group_by(day_of_week) |>
  summarize(
    avg_accidents = mean(total),
    median_accidents = median(total),
    max_accidents = max(total),
    min_accidents = min(total)
  )

# Display day of week summary
day_of_week_summary |> kable(digits = 2)

# Visualize by day of week
ggplot(accidents, aes(x = day_of_week, y = total)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Accident Distribution by Day of Week",
       x = "Day of Week",
       y = "Number of Accidents",
       caption = "Source: Prince George's County Accident Data")
```

### Task 3 (2 points)
Based on the day-of-week analysis, which day(s) have the highest and lowest average accidents? Why might these patterns exist?

ANSWER HERE: Friday seems to have the highest average accidents and Monday seems to have the lowest according the the box plot. My guess may be that on Fridays, there are more people going out with friends/family, or speeding may be higher from people rushing to get back home for the weekend. In general, since Friday is the last day of work for many people, it could be that people tend to be more lax and not as attentive as they were during the start of the week on Monday. 

## Calculating Z-Scores

### Task 4 (2 points)

Now, using the accidents dataframe, add three columns using `mutate()`: mean_accidents, sd_accidents and z_score, based on the example from class.

```{r calculate-z-scores}
# Calculate z-scores for total accidents
 accidents <- accidents |>
  mutate(
    mean_accidents = mean(total),
    sd_accidents = sd(total),
    z_score = (total - mean_accidents) / sd_accidents
    )

# Display the data with z-scores
accidents |>
  select(date, day_of_week, total, z_score) |>
  head(10) |>
  kable(digits = 2)
```

## Identifying Unusual Days

Let's identify days with unusually high or low accident numbers (|z-score| > 2):

```{r unusual-days}
# Find days with unusual accident numbers
unusual_days <- accidents |>
  filter(abs(z_score) > 2) |>
  arrange(desc(abs(z_score)))

# Display number of unusual days
cat("Number of days with unusual accident counts:", nrow(unusual_days), 
    "out of", nrow(accidents), "total days (", 
    round(nrow(unusual_days)/nrow(accidents)*100, 2), "%)")

# Display the top 10 most unusual days
unusual_days |>
  head(10) |>
  select(date, day_of_week, total, z_score) |>
  kable(digits = 2)
```

### Task 5 (3 points)
Choose one of the top 10 unusual days and research what might have happened on that date. Look for weather events, holidays, major incidents, or other factors that could explain the unusual number of accidents. Briefly describe your findings and how you would incorporate this context into a story.

ANSWER HERE: I picked the date October 29th, 2021, with 58 accidents. On this day, there was a massive, unnamed coastal storm that swept across the mid-atlantic, which trigged a surge in traffic accidents and severe road hazards across Prince Georges County and the rest of Maryland. I'd incorporate this into a story by emphasizing how the coastal storm led to an abnormally high rate of accidents compared to typical accident numbers in the County.

Let's visualize these unusual days:

```{r plot-unusual-days}
# Highlight unusual days in the time series
ggplot(accidents, aes(x = date, y = total, color = abs(z_score) > 2)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("gray", "red"), 
                     name = "Unusual Value",
                     labels = c("Normal", "Unusual")) +
  theme_minimal() +
  labs(title = "Daily Accidents with Unusual Days Highlighted",
       subtitle = "Unusual days have z-scores with absolute values > 2",
       x = "Date",
       y = "Number of Accidents",
       caption = "Source: Prince George's County Accident Data")
```

## Day of Week Patterns

### Task 6 (3 points)

Let's calculate z-scores that account for day of week patterns based on the same pattern as above:

```{r day-of-week-z-scores}
# Calculate day-of-week adjusted z-scores
accidents <- accidents |>
  group_by(day_of_week) |>
  mutate(
    dow_mean = mean(total),
    dow_sd = sd(total),
    dow_z_score = (total- dow_mean) / dow_sd
  ) |>
  ungroup()

# Find days with unusual day-of-week z-scores
unusual_dow <- accidents |>
  filter(abs(dow_z_score) > 2) |>
  arrange(desc(abs(dow_z_score)))

# Display the top unusual days adjusted for day of week
unusual_dow |>
  head(10) |>
  select(date, day_of_week, total, dow_mean, dow_z_score) |>
  kable(digits = 2)
```

### Task 7 (3 points)
Compare the days that have extreme z-scores overall versus those with extreme day-of-week z-scores. How do these lists differ? What journalistic insights can you draw from comparing these two different approaches to measuring "unusual" days?

ANSWER HERE:Fridays and Saturdays stand out to me as they average about 28 accidents verses roughly 23 on Tuesday through Thursday. I know that day of the week z scored compare each day to its own weekday, wheras overall z scores compare everyday to the dataset average. Journalistically, the methods answer different questions. Overall z-scores show when risk is the highest which could matter for staffing and public warnings, but they mostly confirm that weekends are riskier. Friday 10/29/2021 is extreme by both measures which means its the strongest story. 

Let's see how many unusual days we have for each day of the week:

```{r unusual-by-dow}
# Count unusual days by day of week
unusual_count_by_dow <- accidents |>
  filter(abs(dow_z_score) > 2) |>
  count(day_of_week) |>
  arrange(desc(n))

# Get total count by day of week for comparison
total_by_dow <- accidents |>
  count(day_of_week)

# Join to calculate percentages
unusual_rates <- unusual_count_by_dow |>
  left_join(total_by_dow, by = "day_of_week") |>
  rename(unusual_days = n.x, total_days = n.y) |>
  mutate(
    pct_unusual = (unusual_days / total_days) * 100
  )

# Display results
unusual_rates |> kable(digits = 2)
```

## Monthly Patterns

Let's examine if certain months have more unusual days:

```{r monthly-patterns}
# Extract month from date for seasonal analysis
accidents <- accidents |>
  mutate(month_of_year = month(date, label = TRUE))

# Calculate average accidents by month
monthly_pattern <- accidents |>
  group_by(month_of_year) |>
  summarize(
    avg_accidents = mean(total),
    median_accidents = median(total)
  )

# Plot monthly pattern
ggplot(monthly_pattern, aes(x = month_of_year, y = avg_accidents, group = 1)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Average Daily Accidents by Month",
       x = "Month",
       y = "Average Number of Accidents",
       caption = "Source: Prince George's County Accident Data")

# Calculate month-adjusted z-scores
accidents <- accidents |>
  group_by(month_of_year) |>
  mutate(
    monthly_mean = mean(total),
    monthly_sd = sd(total),
    monthly_z_score = (total - monthly_mean) / monthly_sd
  ) |>
  ungroup()

# Find days with unusual monthly z-scores
unusual_monthly <- accidents |>
  filter(abs(monthly_z_score) > 2) |>
  arrange(desc(abs(monthly_z_score)))

# Display top days that are unusual for their month
unusual_monthly |>
  head(10) |>
  select(date, month_of_year, day_of_week, total, monthly_mean, monthly_z_score) |>
  kable(digits = 2)
```

### Task 8 (2 points)
Based on the monthly pattern analysis, which months have the highest and lowest average accidents? Develop a hypothesis about why these seasonal patterns exist.

ANSWER HERE: January has the lowest average accidents and October has the highest average, this could be because there is less driving overall after the holidays, with fewer trips, events, and nights out. Bad weather keeps people home, and those who do drive tend to go slower and more carefully. October being the highest could be because school and work are back in sesson 

## Identifying Extreme Outliers

Let's look specifically at days with extremely unusual accident counts (z-scores beyond ±3):

```{r extreme-outliers}
# Find extreme outliers
extreme_outliers <- accidents |>
  filter(abs(z_score) > 3) |>
  arrange(desc(abs(z_score)))

# Display extreme outliers
if(nrow(extreme_outliers) > 0) {
  extreme_outliers |>
    select(date, day_of_week, total, z_score) |>
    kable(digits = 2)
} else {
  cat("No days with z-scores beyond ±3 were found in the dataset.")
}
```

### Task 9 (2 points)
For the extreme outliers identified (z-scores beyond ±3), what do they have in common?

ANSWER HERE: Most extreme outliers are high accident Fridays and Saturdays, concentrated in 2021, with suspiciously low days that likely reflect missing data. Most outliers fall in spring and summer months or near late fall or holidays.

## Your Best Story Pitch

### Task 10 (5 points)
Based on all the analyses performed, craft your best story pitch, including a headline and lede. Be sure to include what kinds of sources/people you'd talk to for the story.

Pitch: beyond the weekend rush, the ordinary days when crashes spiked

Headline: On Normal weekdays, crashes doubled. No one saw it coming. 

Lede: On Friday, Oct. 29, 2021, Prince Georges County recorded 58 accidents, more than double the typical day. But it wasn't the only the day that defied the pattern. On ordinary Tuesdays, Wednesdays, and Thursdays, when the area usually sees about 23 crashes, six days saw 42 to 45. 

Sources I'd talk to: Local police or state highway patrol: ask them what happened on the spike dates, and whether any reporting changes explain the accounts
State or county transportation department: I'd ask for roach closures, construction, detours, and signal outages
National weather service records or metereologists: weather can highly impact accidents so it would be important to get weather perspective for these days 
