---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
mandms <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset


```{r}
sample_5 <- mandms |> sample_n(5)
```

### Task 4: Calculate the mean age and standard deviation of the sample, using either the red or black number of M&Ms
```{r}
set.seed(42)
sample_5|>
  summarize(
    mean_red = mean(red),
    sd_red = sd(red),
 )
```


### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4

```{r}
set.seed(42)
mandms |>
  summarize(
    mean_red = mean(red),
    sd_red = sd(red),
  )
```


### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar? 


They're actually very similar, the biggest jump in raw numbers would be the mean that went from 76 to 72.98 which makes sense because we used a small sample size. 

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

Even though I only used five rows of data for the sample, the sample was still pretty similar to the full dataset overall. The mean was only a few points different and the sd was close as well. I think this shows that random sampling can still give you a good idea of a larger dataset, even when you're only looking at a small sample out of that data. 


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
